---
title: "dataWrangling"
output: html_document
---

```{r setup, include=FALSE}
# include packages
suppressPackageStartupMessages(library(shinydashboard))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(shiny))
suppressPackageStartupMessages(library(shinyjs))
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(htmltools))
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# read in the csv file
delaydat = read_csv("https://raw.githubusercontent.com/ytai9109/AirDashboard/main/848649355_12021_1941_airline_delay_causes.csv")

# round carrier_ct, weather_ct, nas_ct, security_ct, late_aircraft_ct to whole numbers
delaydat = delaydat %>% 
    mutate(across(9:13, round, digits=0))
```

- group data for total arrival and delayed flights by month
- sum of delayed flights consist of delayed + cancelled + diverted
- take natural log of arrival and delay counts
- demean the arrival and delay counts
```{r}
dat1 = delaydat %>%
    mutate(month = month.abb[month]) %>%
    unite(year, month, col = "year_month", sep="/", remove = FALSE) %>%
    group_by(year,month) %>%
    mutate(total_arrival_flights = sum(arr_flights, na.rm = TRUE)) %>%
    mutate(total_delayed_flights = sum(arr_del15, arr_cancelled, arr_diverted, na.rm = TRUE)) %>%
    mutate(percent_delayed = total_delayed_flights / total_arrival_flights) %>%
    distinct(year_month, .keep_all = TRUE) %>%
    ungroup() %>%
    select(year_month, year, month, total_arrival_flights, total_delayed_flights, percent_delayed) %>%
    # take ln of arrival and delay flight count
    mutate(ln_total_arrival_flights = log(total_arrival_flights)) %>% 
    mutate(ln_total_arrival_flights = ifelse(ln_total_arrival_flights < 0, 0, ln_total_arrival_flights)) %>%
    mutate(ln_total_delayed_flights = log(total_delayed_flights)) %>% 
    mutate(ln_total_delayed_flights = ifelse(ln_total_delayed_flights < 0, 0, ln_total_delayed_flights)) %>%
    mutate(demean_total_arrival_flights = round(total_arrival_flights - mean(total_arrival_flights), digits = 0)) %>%
    mutate(demean_total_delayed_flights = round(total_delayed_flights - mean(total_delayed_flights), digits = 0)) %>%
    group_by(year_month, year, month) %>%
    mutate(month = factor(month,levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>%
    arrange(year, month)
```


```{r}
dat1 %>% filter(year %in% c(2004,2009))
```


build the plot
- need to make y data dynamic (arrival count, delay count, percent delayed)
- need to make y-axis scale dynamic
- need to make title dynamic
- need to make ylabel dynamic
```{r}
g1 = ggplot(data = dat1 %>% filter(year %in% c(2004,2009)), aes(x = year_month, y = percent_delayed, text = paste0("<b>Year/Month:</b> ", .data[["year_month"]], "\n<b>Total arrival count:</b> ", .data[["total_arrival_flights"]], "\n<b>Total delayed count:</b> ", .data[["total_delayed_flights"]], "\n<b>Percentage delayed:</b> ", round(.data[["percent_delayed"]]*100, digits = 2), "%"))) +
    geom_bar(stat = "identity", 
             fill = "steelblue") +
    geom_smooth(aes(group = 1), method = "loess", span = 0.4, SE = FALSE, fill = NA) + 
    scale_x_discrete(limits = dat1 %>% filter(year %in% c(2004,2009)) %>% pull(`year_month`)) +
    labs(title = "Percent of Delayed Flights in 2004",
         x = "Months",
         y = "Percent Delayed") +
    theme(axis.text.x = element_text(angle = 60, hjust=1, vjust=0.2)) +
        theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold", margin = margin(t = 0, r = 0, b = 10, l = 0))) +
        theme(axis.title.x = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), face = "bold")) +
        theme(axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), face = "bold"))

g1 = ggplotly(g1, tooltip = "text")

g1
```



```{r}
g1 = ggplot(data = dat1 %>% filter(year %in% c(2019, 2020)), aes(x = year_month, y = percent_delayed, text = paste0("<b>Year/Month:</b> ", .data[["year_month"]], "\n<b>Total arrival count:</b> ", .data[["total_arrival_flights"]], "\n<b>Total delayed count:</b> ", .data[["total_delayed_flights"]], "\n<b>Percentage delayed:</b> ", round(.data[["percent_delayed"]]*100, digits = 2), "%"))) +
    geom_line(aes(group = 1)) +
    geom_point() +
    geom_smooth(aes(group = 1), method = "lm", SE = FALSE, fill = NA) +
    scale_x_discrete(limits = dat1 %>% filter(year %in% c(2019,2020)) %>% pull(`year_month`)) +
    labs(title = "Percent of Delayed Flights in 2004",
         x = "Months",
         y = "Percent Delayed") +
    theme(axis.text.x = element_text(angle = 60, hjust=1, vjust=0.2)) +
        theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold", margin = margin(t = 0, r = 0, b = 10, l = 0))) +
        theme(axis.title.x = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), face = "bold")) +
        theme(axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), face = "bold"))


g1
```

```{r}
dat1_rowsize = dat1 %>% filter(year %in% c(2019, 2020)) %>% nrow()

dat1_fit = dat1 %>% 
    filter(year %in% c(2019, 2020)) %>%
    add_column(count = c(1:dat1_rowsize)) %>%
    relocate(count, .before = year_month)
```


```{r}
fit = lm(total_arrival_flights ~ count, data = dat1_fit)

new_count = data.frame(count = c((dat1_rowsize+1): 30))

predict_output = as.data.frame(predict(fit, new_count))
months_predicted = c(1: (length(c((dat1_rowsize+1): 30))))
predict_output = cbind(months_predicted, predict_output)
names(predict_output)[1] = "Months Ahead"
names(predict_output)[2] = "Predicted Value"

```


####################################

GOAL: number of delays for an airport at a given time

```{r message=FALSE, warning=FALSE, include=FALSE}
# read in the csv file
delaydat = read_csv("848649355_12021_1941_airline_delay_causes.csv")

# round carrier_ct, weather_ct, nas_ct, security_ct, late_aircraft_ct to whole numbers
delaydat = delaydat %>% 
    mutate(across(9:13, round, digits=0))

delay_States = delaydat %>%
    #group into US Regions
    mutate(region = ifelse(grepl("AK:|HI:|WA:|OR:|ID:|MT:|WY:|CA:|NV:|UT:|CO:", airport_name), "West", "-")) %>%
    mutate(region = ifelse(grepl("ND:|MN:|WI:|MI:|SD:|IA:|NE:|KS:|MO:|IL:|IN:|OH:", airport_name), "Midwest", region)) %>%
    mutate(region = ifelse(grepl("AZ:|NM:|TX:|OK:", airport_name), "Southwest", region)) %>%
    mutate(region = ifelse(grepl("PA:|NY:|VT:|ME:|NH:|MA:|RI:|CT:|NJ:", airport_name), "Northeast", region)) %>%
    mutate(region = ifelse(grepl("AR:|LA:|MS:|AL:|GA:|FL:|KY:|TN:|MD:|DE:|WV:|DC:|VA:|NC:|SC:", airport_name), "Southeast", region)) %>%
    mutate(region = ifelse(grepl("West|Midwest|Southwest|Northeast|Southeast",region), region, "Outer")) %>%
    relocate(region, .after = month) %>%
    # group into US states (there are 52 states.....)
    mutate(state_abv = ifelse(grepl("AL:", airport_name), "AL", "-")) %>%
    mutate(state_full = ifelse(grepl("AL:", airport_name), "Alabama", "-")) %>%
    mutate(state_abv = ifelse(grepl("AK:", airport_name), "AK", state_abv)) %>%
    mutate(state_full = ifelse(grepl("AK:", airport_name), "Alaska", state_full)) %>%
    mutate(state_abv = ifelse(grepl("AS:", airport_name), "AS", state_abv)) %>%
    mutate(state_full = ifelse(grepl("AS:", airport_name), "American Samoa", state_full)) %>%
    mutate(state_abv = ifelse(grepl("AZ:", airport_name), "AZ", state_abv)) %>%
    mutate(state_full = ifelse(grepl("AZ:", airport_name), "Arizona", state_full)) %>%
    mutate(state_abv = ifelse(grepl("AR:", airport_name), "AR", state_abv)) %>%
    mutate(state_full = ifelse(grepl("AR:", airport_name), "Arkansas", state_full)) %>%
    mutate(state_abv = ifelse(grepl("CA:", airport_name), "CA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("CA:", airport_name), "California", state_full)) %>%
    mutate(state_abv = ifelse(grepl("CO:", airport_name), "CO", state_abv)) %>%
    mutate(state_full = ifelse(grepl("CO:", airport_name), "Colorado", state_full)) %>%
    mutate(state_abv = ifelse(grepl("CT:", airport_name), "CT", state_abv)) %>%
    mutate(state_full = ifelse(grepl("CT:", airport_name), "Connecticut", state_full)) %>%
    mutate(state_abv = ifelse(grepl("DE:", airport_name), "DE", state_abv)) %>%
    mutate(state_full = ifelse(grepl("DE:", airport_name), "Delaware", state_full)) %>%
    mutate(state_abv = ifelse(grepl("DC:", airport_name), "DC", state_abv)) %>%
    mutate(state_full = ifelse(grepl("DC:", airport_name), "District of Columbia", state_full)) %>%
    mutate(state_abv = ifelse(grepl("FL:", airport_name), "FL", state_abv)) %>%
    mutate(state_full = ifelse(grepl("FL:", airport_name), "Florida", state_full)) %>%
    mutate(state_abv = ifelse(grepl("GA:", airport_name), "GA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("GA:", airport_name), "Georgia", state_full)) %>%
    mutate(state_abv = ifelse(grepl("GU:", airport_name), "GU", state_abv)) %>%
    mutate(state_full = ifelse(grepl("GU:", airport_name), "Guam", state_full)) %>%
    mutate(state_abv = ifelse(grepl("HI:", airport_name), "HI", state_abv)) %>%
    mutate(state_full = ifelse(grepl("HI:", airport_name), "Hawaii", state_full)) %>%
    mutate(state_abv = ifelse(grepl("ID:", airport_name), "ID", state_abv)) %>%
    mutate(state_full = ifelse(grepl("ID:", airport_name), "Idaho", state_full)) %>%
    mutate(state_abv = ifelse(grepl("IL:", airport_name), "IL", state_abv)) %>%
    mutate(state_full = ifelse(grepl("IL:", airport_name), "Illinois", state_full)) %>%
    mutate(state_abv = ifelse(grepl("IN:", airport_name), "IN", state_abv)) %>%
    mutate(state_full = ifelse(grepl("IN:", airport_name), "Indiana", state_full)) %>%
    mutate(state_abv = ifelse(grepl("IA:", airport_name), "IA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("IA:", airport_name), "Iowa", state_full)) %>%
    mutate(state_abv = ifelse(grepl("KS:", airport_name), "KS", state_abv)) %>%
    mutate(state_full = ifelse(grepl("KS:", airport_name), "Kansas", state_full)) %>%
    mutate(state_abv = ifelse(grepl("KY:", airport_name), "KY", state_abv)) %>%
    mutate(state_full = ifelse(grepl("KY:", airport_name), "Kentucky", state_full)) %>%
    mutate(state_abv = ifelse(grepl("LA:", airport_name), "LA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("LA:", airport_name), "Louisiana", state_full)) %>%
    mutate(state_abv = ifelse(grepl("ME:", airport_name), "ME", state_abv)) %>%
    mutate(state_full = ifelse(grepl("ME:", airport_name), "Maine", state_full)) %>%
    mutate(state_abv = ifelse(grepl("MD:", airport_name), "MD", state_abv)) %>%
    mutate(state_full = ifelse(grepl("MD:", airport_name), "Maryland", state_full)) %>%
    mutate(state_abv = ifelse(grepl("MA:", airport_name), "MA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("MA:", airport_name), "Massachusetts", state_full)) %>%
    mutate(state_abv = ifelse(grepl("MI:", airport_name), "MI", state_abv)) %>%
    mutate(state_full = ifelse(grepl("MI:", airport_name), "Michigan", state_full)) %>%
    mutate(state_abv = ifelse(grepl("MN:", airport_name), "MN", state_abv)) %>%
    mutate(state_full = ifelse(grepl("MN:", airport_name), "Minnesota", state_full)) %>%
    mutate(state_abv = ifelse(grepl("MS:", airport_name), "MS", state_abv)) %>%
    mutate(state_full = ifelse(grepl("MS:", airport_name), "Mississippi", state_full)) %>%
    mutate(state_abv = ifelse(grepl("MO:", airport_name), "MO", state_abv)) %>%
    mutate(state_full = ifelse(grepl("MO:", airport_name), "Missouri", state_full)) %>%
    mutate(state_abv = ifelse(grepl("MT:", airport_name), "MT", state_abv)) %>%
    mutate(state_full = ifelse(grepl("MT:", airport_name), "Montana", state_full)) %>%
    mutate(state_abv = ifelse(grepl("NE:", airport_name), "NE", state_abv)) %>%
    mutate(state_full = ifelse(grepl("NE:", airport_name), "Nebraska", state_full)) %>%
    mutate(state_abv = ifelse(grepl("NV:", airport_name), "NV", state_abv)) %>%
    mutate(state_full = ifelse(grepl("NV:", airport_name), "Nevada", state_full)) %>%
    mutate(state_abv = ifelse(grepl("NH:", airport_name), "NH", state_abv)) %>%
    mutate(state_full = ifelse(grepl("NH:", airport_name), "New Hampshire", state_full)) %>%
    mutate(state_abv = ifelse(grepl("NJ:", airport_name), "NJ", state_abv)) %>%
    mutate(state_full = ifelse(grepl("NJ:", airport_name), "New Jersey", state_full)) %>%
    mutate(state_abv = ifelse(grepl("NM:", airport_name), "NM", state_abv)) %>%
    mutate(state_full = ifelse(grepl("NM:", airport_name), "New Mexico", state_full)) %>%
    mutate(state_abv = ifelse(grepl("NY:", airport_name), "NY", state_abv)) %>%
    mutate(state_full = ifelse(grepl("NY:", airport_name), "New York", state_full)) %>%
    mutate(state_abv = ifelse(grepl("NC:", airport_name), "NC", state_abv)) %>%
    mutate(state_full = ifelse(grepl("NC:", airport_name), "North Carolina", state_full)) %>%
    mutate(state_abv = ifelse(grepl("ND:", airport_name), "ND", state_abv)) %>%
    mutate(state_full = ifelse(grepl("ND:", airport_name), "North Dakota", state_full)) %>%
    mutate(state_abv = ifelse(grepl("OH:", airport_name), "OH", state_abv)) %>%
    mutate(state_full = ifelse(grepl("OH:", airport_name), "Ohio", state_full)) %>%
    mutate(state_abv = ifelse(grepl("OK:", airport_name), "OK", state_abv)) %>%
    mutate(state_full = ifelse(grepl("OK:", airport_name), "Oklahoma", state_full)) %>%
    mutate(state_abv = ifelse(grepl("OR:", airport_name), "OR", state_abv)) %>%
    mutate(state_full = ifelse(grepl("OR:", airport_name), "Oregon", state_full)) %>%
    mutate(state_abv = ifelse(grepl("PA:", airport_name), "PA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("PA:", airport_name), "Pennsylvania", state_full)) %>%
    mutate(state_abv = ifelse(grepl("PR:", airport_name), "PR", state_abv)) %>%
    mutate(state_full = ifelse(grepl("PR:", airport_name), "Puerto Rico", state_full)) %>%
    mutate(state_abv = ifelse(grepl("RI:", airport_name), "RI", state_abv)) %>%
    mutate(state_full = ifelse(grepl("RI:", airport_name), "Rhode Island", state_full)) %>%
    mutate(state_abv = ifelse(grepl("SC:", airport_name), "SC", state_abv)) %>%
    mutate(state_full = ifelse(grepl("SC:", airport_name), "South Carolina", state_full)) %>%
    mutate(state_abv = ifelse(grepl("SD:", airport_name), "SD", state_abv)) %>%
    mutate(state_full = ifelse(grepl("SD:", airport_name), "South Dakota", state_full)) %>%
    mutate(state_abv = ifelse(grepl("TN:", airport_name), "TN", state_abv)) %>%
    mutate(state_full = ifelse(grepl("TN:", airport_name), "Tennessee", state_full)) %>%
    mutate(state_abv = ifelse(grepl("TX:", airport_name), "TX", state_abv)) %>%
    mutate(state_full = ifelse(grepl("TX:", airport_name), "Texas", state_full)) %>%
    mutate(state_abv = ifelse(grepl("UT:", airport_name), "UT", state_abv)) %>%
    mutate(state_full = ifelse(grepl("UT:", airport_name), "Utah", state_full)) %>%
    mutate(state_abv = ifelse(grepl("VT:", airport_name), "VT", state_abv)) %>%
    mutate(state_full = ifelse(grepl("VT:", airport_name), "Vermont", state_full)) %>%
    mutate(state_abv = ifelse(grepl("VA:", airport_name), "VA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("VA:", airport_name), "Virginia", state_full)) %>%
    mutate(state_abv = ifelse(grepl("VI:", airport_name), "VI", state_abv)) %>%
    mutate(state_full = ifelse(grepl("VI:", airport_name), "Virgin Islands", state_full)) %>%
    mutate(state_abv = ifelse(grepl("WA:", airport_name), "WA", state_abv)) %>%
    mutate(state_full = ifelse(grepl("WA:", airport_name), "Washington", state_full)) %>%
    mutate(state_abv = ifelse(grepl("WV:", airport_name), "WV", state_abv)) %>%
    mutate(state_full = ifelse(grepl("WV:", airport_name), "West Virginia", state_full)) %>%
    mutate(state_abv = ifelse(grepl("WI:", airport_name), "WI", state_abv)) %>%
    mutate(state_full = ifelse(grepl("WI:", airport_name), "Wisconsin", state_full)) %>%
    mutate(state_abv = ifelse(grepl("WY:", airport_name), "WY", state_abv)) %>%
    mutate(state_full = ifelse(grepl("WY:", airport_name), "Wyoming", state_full)) %>%
    mutate(state_abv = ifelse(grepl("TT:", airport_name), "TT", state_abv)) %>%
    mutate(state_full = ifelse(grepl("TT:", airport_name), "Trust Territory of the Pacific Islands", state_full)) %>%
    relocate(state_abv, .after = region) %>%
    relocate(state_full, .after = state_abv)
```


see what kind of airports there are
416 airports available, looks like I need a selector widget... (ones that search by typing)
```{r}
delaydat %>% select(airport, airport_name) %>%
    distinct(airport, airport_name) %>%
    arrange(airport)
```
- make airport selection dynamic
- make time selection dynamic
-- if one month selected, then display bar plot
-- if multiple months selected, then display side by side bar plot
-- by default, display 6 months

**test
```{r message=FALSE, warning=FALSE}
dat2_delay_count = delaydat %>% 
    group_by(year, month, airport) %>%
    summarize(delayed_count = sum(arr_del15, arr_cancelled, arr_diverted, na.rm = TRUE)) 
```
**test
```{r}
dat2_carrier_count = delaydat %>%
    group_by(year, month, airport) %>%
    mutate(carrier_count = sum(carrier_ct, na.rm = TRUE)) %>%
    arrange(year, month, airport) %>%
    distinct(year, month, airport, .keep_all = TRUE)
```


- find the count for each mode of delay for each airport, for a specific year and month
- arrange by airport names
- get rid of airlines
- create full airport name
- create year/month column
- sort year/month column by month factors
- make into "long" data frame for plotting
- calculate percentage for a specific delay cause out of all delays for that specific airport for a given year and month
```{r}
dat2 = delay_States %>% 
    group_by(year, month, airport) %>%
    mutate(delayed_count = sum(arr_del15, arr_cancelled, arr_diverted, na.rm = TRUE)) %>%
    mutate(carrier_count = sum(carrier_ct, na.rm = TRUE)) %>%
    mutate(weather_count = sum(weather_ct, na.rm = TRUE)) %>%
    mutate(nas_count = sum(nas_ct, na.rm = TRUE)) %>%
    mutate(security_count = sum(security_ct, na.rm = TRUE)) %>%
    mutate(late_aircraft_count = sum(late_aircraft_ct, na.rm = TRUE)) %>%
    mutate(canceled_count = sum(arr_cancelled, na.rm = TRUE)) %>%
    mutate(diverted_count = sum(arr_diverted, na.rm = TRUE)) %>%
    mutate(arrival_count = sum(arr_flights, na.rm = TRUE)) %>%
    arrange(year, month, airport) %>%
    distinct(year, month, airport, .keep_all = TRUE) %>%
    select(year, month, region, state_abv, state_full, airport, airport_name, 
           arrival_count, delayed_count, carrier_count, weather_count, 
           nas_count, security_count, late_aircraft_count, 
           canceled_count, diverted_count) %>%
    unite(airport, airport_name, 
          col = "full_airport_name", 
          sep = " | ", 
          remove = FALSE) %>%
    ungroup() %>%
    select(-airport, -airport_name) %>%
    mutate(month = month.abb[month]) %>%
    unite(year, month, col = "year_month", sep="/", remove = FALSE) %>%
    mutate(month = factor(month,levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>%
    pivot_longer(cols = carrier_count:diverted_count, 
                 names_to = "delay_types", 
                 values_to = "count") %>%
    mutate(count_percent = ifelse(delayed_count == 0 & count == 0, 0, count/delayed_count)) %>%
    mutate(count_percent = round(count_percent * 100, digits = 2))
    
    
```

```{r}
list_of_yearmonths = dat2 %>% distinct(year_month) %>% pull(year_month)
list_of_airports = dat2 %>% distinct(full_airport_name) %>% pull(full_airport_name)
list_of_states = delay_States %>% distinct(state_full) %>% arrange(state_full) %>% pull(state_full)
```


- create dynamic bar plot of delay causes for each airport for a given year and month
```{r, width = 400}
g2 = ggplot(data = dat2 %>% 
                filter(full_airport_name == "AEX | Alexandria, LA: Alexandria International") %>% 
                filter(year == "2010" & month %in% c("Jun", "Jul")), 
            aes(x = delay_types, y = count_percent, fill = year_month)) +
    geom_bar(stat = "identity", position='dodge') + 
    coord_flip() +
    scale_x_discrete(labels = rev(c("Carrier", "Weather", "NAS", "Security", "Late Aircraft", "Canceled", "Diverted")), limits = rev(c("carrier_count", "weather_count", "nas_count", "security_count", "late_aircraft_count", "canceled_count", "diverted_count"))) +
    labs(title = "Delay causes for airport",
         x = "Delay Cause",
         y = "Count") +
    theme(axis.text.x = element_text(angle = 60, hjust=1, vjust=0.2)) +
    scale_fill_discrete(name = "Year/Month") +
    theme(plot.title = element_text(size = 14, hjust = 0.5, face = "bold", margin = margin(t = 0, r = 0, b = 10, l = 0))) +
    theme(axis.title.x = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), face = "bold")) +
    theme(axis.title.y = element_text(margin = margin(t = 0, r = 0, b = 0, l = 0), face = "bold")) +
    theme(plot.margin = unit(c(1.8,0.5,0.5,0.5), "cm"))
    
g2 = ggplotly(g2, tooltip = c("x","y","fill"))

for (i in 1:length(g2$x$data)){
    if (!is.null(g2$x$data[[i]]$name)){
        g2$x$data[[i]]$name = dat2 %>% 
                filter(full_airport_name == "AEX | Alexandria, LA: Alexandria International") %>% 
                filter(year == "2010" & month %in% c("Jun", "Jul")) %>%
            distinct(year_month) %>% 
            pull(year_month) %>% 
            nth(i)
    }
}

g2
```


- for pie charts I will have to use plotly

```{r}
# rename delay types for pie chart
dat2_pie = dat2 %>%
    mutate(delay_types = replace(delay_types, delay_types == "carrier_count", "Carrier")) %>%
    mutate(delay_types = replace(delay_types, delay_types == "weather_count", "Weather")) %>%
    mutate(delay_types = replace(delay_types, delay_types == "nas_count", "NAS")) %>%
    mutate(delay_types = replace(delay_types, delay_types == "security_count", "Security")) %>%
    mutate(delay_types = replace(delay_types, delay_types == "late_aircraft_count", "Late Aircraft")) %>%
    mutate(delay_types = replace(delay_types, delay_types == "canceled_count", "Canceled")) %>%
    mutate(delay_types = replace(delay_types, delay_types == "diverted_count", "Diverted"))
```


```{r}
g3 = plot_ly(data = dat2_pie %>% filter(full_airport_name == "AEX | Alexandria, LA: Alexandria International") %>% filter(year == "2010" & month %in% c("Jun")),
             labels = ~delay_types,
             values = ~count,
             type = "pie",
             textposition = 'outside',
             automargin = TRUE,
             texttemplate = "<b>%{label}</b>\n%{percent}",
             insidetextfont = list(color = '#FFFFFF'),
             hovertemplate = "<b>%{label}</b>\nCount: %{value}<extra></extra>",
             marker = list(line = list(color = '#FFFFFF', width = 1.5)),
             showlegend = TRUE) %>%
    layout(title = list(text = 'Delay Types: Aiport in (Year)(Month)',
                        xanchor = "center"),
           xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
           yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
           # default padding is 80px
           margin = list(l = 80, r = 80, t = 120, b = 20, pad = 0),
           legend = list(borderwidth = 2, 
                         itemclick = FALSE, 
                         itemdoubleclick = FALSE, 
                         title = list(text = "<b>Delay Types<b>")))

g3
```


- group airports into US regions, search by state abbreviations
```{r}
dat3 = dat2 %>%
    rename(delay_type_percent = count_percent) %>%
    rename(delay_type_count = count) %>%
    mutate(delay_percent = delayed_count / arrival_count) %>%
    relocate(delay_percent, .after = delayed_count)
    
```


- create scatter plot of percent of delayed flights vs percent of 'cause' causing the delay, scatter point size by arrival size, 
```{r}
g4 = ggplot(data = dat3 %>% filter(year == "2010" & month %in% c("Jun")) %>% filter(delay_types == "carrier_count"), aes(x = delay_type_percent, y = delay_percent)) + 
    geom_point(aes(color = region, 
                   size = arrival_count,
                   text = paste0(full_airport_name, "\n<b>No. of arrivals:</b> ", arrival_count, "\n<b>No. of delays:</b> ", delayed_count ,"\n<b>Percent of total flights delayed:</b> ", round(delay_percent*100, digits = 2), "%\n<b>No. of Carrier delays:</b> ", delay_type_count ,"\n<b>Percent of Carrier delays:</b> ", round(delay_type_percent*100, digits = 2), "%") )) +
    scale_x_continuous(labels = function(x) paste0(x*100, '%')) +
    scale_y_continuous(labels = function(y) paste0(y*100, '%')) +
    labs(x = 'Likelihood carrier is causing delay', 
         y = 'Percentage of delayed flights',
         color = 'Region', 
         size = 'No. of Arrivals',
         title = 'Percentage of delayed flights vs likelihood of weather causing the delay') + 
    guides(size = FALSE) +
    theme(legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) +
    theme(plot.title = element_text(size = 11, hjust = 0.5, face = "bold", margin = margin(t = 0, r = 0, b = 10, l = 0))) +
    theme(axis.title.x = element_text(face = "bold")) +
    theme(axis.title.y = element_text(face = "bold")) +
    theme(plot.margin = unit(c(1,0.5,0.5,0.5), "cm"))
    
g4 = ggplotly(g4, tooltip = "text")

g4
```

- wrangle data for box plot
```{r}
dat4 = delay_States %>% 
    group_by(year, month, carrier) %>%
    mutate(delayed_count = sum(arr_del15, arr_cancelled, arr_diverted, na.rm = TRUE)) %>%
    mutate(Carrier = sum(carrier_ct, na.rm = TRUE)) %>%
    mutate(Weather = sum(weather_ct, na.rm = TRUE)) %>%
    mutate(NAS = sum(nas_ct, na.rm = TRUE)) %>%
    mutate(Security = sum(security_ct, na.rm = TRUE)) %>%
    mutate(`Late Aircraft` = sum(late_aircraft_ct, na.rm = TRUE)) %>%
    mutate(Canceled = sum(arr_cancelled, na.rm = TRUE)) %>%
    mutate(Diverted = sum(arr_diverted, na.rm = TRUE)) %>%
    mutate(arrival_count = sum(arr_flights, na.rm = TRUE)) %>%
    ungroup() %>%
    arrange(year, month, carrier) %>%
    select(year, month, state_abv, state_full, carrier, 
           carrier_name, airport, airport_name, 
           arrival_count, delayed_count, Carrier, Weather, 
           NAS, Security, `Late Aircraft`, 
           Canceled, Diverted) %>%
    # merge airport names together
    unite(airport, airport_name,
          col = "full_airport_name",
          sep = " | ",
          remove = TRUE) %>%
    # merge carrier names together
    unite(`carrier`, carrier_name,
          col = "full_carrier_name",
          sep = " | ",
          remove = FALSE) %>%
    # wrangle year/month
    mutate(month = month.abb[month]) %>%
    unite(year, month, col = "year_month", sep="/", remove = FALSE) %>%
    mutate(month = factor(month,levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))) %>%
    # turn wide data to long data
    pivot_longer(cols = Carrier:Diverted,
                 names_to = "delay_types",
                 values_to = "count") %>%
    # percentage for each delay type
    mutate(count_percent = ifelse(delayed_count == 0 & count == 0, 0, count/delayed_count)) %>%
    # percentage of delays out of all arrival flights
    mutate(delay_percent = ifelse(delayed_count == 0 & arrival_count == 0, 0, delayed_count/arrival_count)) %>%
    relocate(delay_percent, .after = delayed_count) %>%
    # convert to factors for box plots
    mutate(`carrier` = factor(`carrier`))

```


```{r}
list_of_carriers = dat4 %>% distinct(full_carrier_name) %>% pull(full_carrier_name)
```

```{r}
g5 = ggplot(data = dat4 %>% 
                filter(carrier %in% c("AA", "AS", "B6")) %>% 
                group_by(carrier) %>% 
                distinct(year_month, .keep_all = TRUE), 
            aes(x = carrier, y = delayed_count, fill = carrier)) + 
    geom_boxplot() +
    geom_jitter(aes(text = paste0("Carrier: ", .data[["carrier"]])),
                size = 1, 
                stroke = 0.2, 
                shape=16, 
                alpha = 0.7, 
                position=position_jitter(0.1)) +
    labs(x = 'Airline Carrier', 
         y = 'Percentage of delayed flights',
         title = 'Boxplot of Delay Percentage vs Airline Carrier from Jun/2013 to Jan/2021') +
    theme(plot.title = element_text(size = 11, hjust = 0.5, face = "bold", margin = margin(t = 0, r = 0, b = 10, l = 0))) +
    theme(axis.title.x = element_text(face = "bold")) +
    theme(axis.title.y = element_text(face = "bold")) +
    scale_fill_brewer(palette="Dark2") +
    theme(plot.margin = unit(c(1,0.5,0.5,0.5), "cm"))

g5 = ggplotly(g5, tooltip = "text")

g5
```


```{r}
airdat = read_csv("airport_locdata.csv")

content = paste0("<b>Yes: </b>", airdat$name, "\n")

leaflet(data = airdat) %>%
    setView(lat = 41.869561, lng = -120.035524, zoom = 3) %>%
    addTiles(urlTemplate = "https://tiles.stadiamaps.com/tiles/alidade_smooth_dark/{z}/{x}/{y}{r}.png") %>%
    addCircleMarkers(lng = ~longitude_deg,
               lat = ~latitude_deg,
               label = ~name,
               labelOptions = labelOptions(textsize = "15px"),
               popup = content,
               radius = 1.5,
               color = "turquoise",
               fillOpacity = 0)
```
wrangle US airport map - longitude and latitude
```{r echo=FALSE, message=FALSE, warning=FALSE}
airdat = read_csv("us-airports.csv")


```

make airport location csv
add unknown airports into airdat
```{r}
airtest2 = airdat %>% filter(iata_code %in% (delaydat %>% distinct(airport) %>% pull(airport))) %>% select(name, latitude_deg, longitude_deg, iata_code) %>%
    add_row(name = "Rafael Hernandez Airport", latitude_deg = 18.4948611, longitude_deg = -67.1294444, iata_code = "BQN") %>%
    add_row(name = "Guam International Airport", latitude_deg = 13.4839507, longitude_deg = 144.7971271, iata_code = "GUM") %>%
    add_row(name = "Panama City Bay County Airport", latitude_deg = 30.212083, longitude_deg = -85.682806, iata_code = "PFN") %>%
    add_row(name = "Tafuna/Pago Pago International Airport", latitude_deg = -14.3316622, longitude_deg = -170.7115031, iata_code = "PPG") %>%
    add_row(name = "Mercedita Airport", latitude_deg = 18.0087806, longitude_deg = -66.5645208, iata_code = "PSE") %>%
    add_row(name = "Luis Munoz Marin International Airport", latitude_deg = 18.4393992, longitude_deg = -66.0021333, iata_code = "SJU") %>%
    add_row(name = "Francisco C Ada/Saipan International Airport", latitude_deg = 15.1202550, longitude_deg = 145.7299841, iata_code = "SPN") %>%
    add_row(name = "Cyril E King Airport", latitude_deg = 18.3373056, longitude_deg = -64.9733333, iata_code = "STT") %>%
    add_row(name = "Henry E Rohlsen Airport", latitude_deg = 17.7015044, longitude_deg = -64.8019429, iata_code = "STX") %>%
    arrange(iata_code) %>%
    write_csv(file = "airport_locdata.csv")
    
```



```{r}
dat5 = delaydat %>%
    group_by(year, airport) %>%
    mutate(arrival_count = sum(arr_flights, na.rm = TRUE)) %>%
    mutate(delayed_count = sum(arr_del15, arr_cancelled, arr_diverted, na.rm = TRUE)) %>%
    mutate(delay_percent = delayed_count / arrival_count) %>%
    distinct(year, airport, .keep_all = TRUE) %>%
    select(year, airport, arrival_count, delayed_count, delay_percent) %>%
    # for each iata_code, join the lat and long into the data frame
    left_join(airtest2, by = c("airport" = "iata_code")) %>%
    select(year, airport, name, latitude_deg, longitude_deg, arrival_count, delayed_count, delay_percent)
    
```
